Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-28 paper/perfmod.tex
write delicate text to clarify the Jouvet and Brinkerhoff approaches

Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.  This and many other mesh/solver details cannot be pursued here.

@@

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but other models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)), and thus some other manner of determining the time step.

@@

Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling \citep{Bueler2021}.  At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula.  Thus an SIA velocity solution requires optimal $O(m)$, $\alpha=0$, flops.

@@

It appears there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.\footnote{The hybrid \cite{Brinkerhoffetal2017} flowline solver is partly implicit but apparently subject to a time-stepping condition when the margin moves.}  For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, and on ice margin determination on a single grid, shows $\beta=1.8$.  The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries.  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, even using realistic, thus irregular, bed topographies.

@@

These comments complete Table \ref{tab:performancemodel}.  All estimates involve a scheme-dependent constant, with especially large values for the implicit schemes.

>>>>

Our assumption of fixed vertical resolution reflects common usage \citep[for example]{Aschwandenetal2019,BrinkerhoffJohnson2015,Hoffmanetal2018,Lengetal2012,
Winkelmannetal2011}, and it permits a rational comparison of asymptotics, but it is not the only possibility.  For example, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Tuminaroetal2016} solvers use 3D refinement, though with various modifications regarding vertical refinement.  This and many other mesh/solver details cannot be pursued here.

@@

Explicit time-stepping with hybrid and higher-order schemes is somewhat better-studied than for Stokes dynamics, especially over horizontal resolutions relevant to whole ice sheets.  Some hybrid schemes apply the pessimistic time step as an adaptive restriction \citep{Winkelmannetal2011}, but other models apparently require the user to choose a time step through trial and error in some circumstances \citep[for example]{Fischleretal2022,Robinsonetal2022}.  The optimistic time step is supported in theory for a certain higher-order DIVA scheme \citep[see equation (52)]{Robinsonetal2022}, but practical Greenland simulations in the same work actually suggest an intermediate power $\Delta t = O(\Delta x^{1.6})$ (Figure 3(a)).

@@

Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling (or solver complexity \citep{Bueler2021}).  At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula.  Thus an SIA velocity solution requires optimal $O(m)$, $\alpha=0$, flops.

@@

The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=1.8$.  (An earlier non-Newton implementation by \cite{JouvetBueler2012} seems not to scale any better.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.  Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012,Bueler2016}.

The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are also solved implicitly.  \cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.  Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$.
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015

@@

These comments complete Table \ref{tab:performancemodel}.  All estimates involve a scheme-dependent constant, one which must be especially large for the implicit schemes.
